---
title: "ADS-506 Final Project - Baggage Complaints"
author: "Team 4: Sowmiya Kanmani Maruthavanan, Ben Ogle, Vicky van der Wagt"
date: "`r Sys.Date()`"
output:
  pdf_document: default
     keep_md: true
  html_document:
    df_print: paged
knitr:
  opts_chunk:
    warning=FALSE
---

## Import relevant libraries

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(dplyr)
library(forecast)
library(zoo)
library(ggplot2)
library(corrplot)
library(plotly)
```

## Pre-Processing

### Import data

```{r}
airlines_df <- read.csv("baggagecomplaints.csv")
head(airlines_df)
summary(airlines_df)
dim(airlines_df)
```

The dataset contains 252 rows and 8 columns

### Check for null values

```{r}
missing <- colSums(is.na(airlines_df))
print(missing)
```

There are no missing values in this dataset. 


### Convert categorical columns into factors

```{r}
#converting month and year to categorical 
airlines_df$Month <- factor(airlines_df$Month)
airlines_df$Year <- factor(airlines_df$Year)
airlines_df$Airline <- factor(airlines_df$Airline)
```


### Transform data into time series

The time series starts in 2004 and ends in 2010. The frequency is set to 12 because the data is monthly and there are 12 months in a year. 

```{r}
# Create time series for 3 airlines
american_eagle.ts <- ts(airlines_df[airlines_df$Airline == "American Eagle", "Baggage"], 
                        start = c(2004,1), end = c(2010,12), freq = 12)

hawaiian.ts <- ts(airlines_df[airlines_df$Airline == "Hawaiian", "Baggage"], 
                        start = c(2004,1), end = c(2010,12), freq = 12)

united.ts <- ts(airlines_df[airlines_df$Airline == "United", "Baggage"], 
                        start = c(2004,1), end = c(2010,12), freq = 12)
```

### Correlation plot

```{r}
numerical <- airlines_df[sapply(airlines_df, is.numeric)]
cor_matrix <- cor(numerical)
corrplot(cor_matrix,
         method = "number",
         tl.col = "black")
```

There are positive correlations between all the numerical variables 'Baggage', 'Scheduled', 'Canceleled', and 'Enplaned'. High correlations exist between 'Baggage' and 'Scheduled', as well as 'Baggage' and 'Enplaned.' Moderate correlations exist between 'Cancelled' and 'Baggage', as well as 'Bagged' and 'Enplaned.' There is a very weak relationship between 'Cancelled' and 'Emplaned.'

### Feature Creation
```{r}
airlines_df$Cancelled_prop <- (airlines_df$Cancelled / airlines_df$Scheduled) * 100
airlines_df$Flights <- airlines_df$Scheduled - airlines_df$Cancelled
airlines_df$Enplaned_per_flight <- airlines_df$Enplaned / airlines_df$Flights
airlines_df$Complaints_per_enplaned <- (airlines_df$Baggage / airlines_df$Enplaned) * 100 

```

#### Created 4 new calculated features:

* Cancelled_prop: the proportion of cancelled flights
* Flights: the number of flights that were not cancelled
* Enplaned_per_flight: the number of enplaned passengers per flight
* Complaints per enplaned: the proportion of complaints over the total number of passengers

### Correlation matrix after feature creation

View the correlation matrix after new features are created

```{r}
numerical <- airlines_df[sapply(airlines_df, is.numeric)]
cor_matrix <- cor(numerical)
corrplot(cor_matrix,
         method = "number",
         tl.col = "black",
         tl.cex = 0.6)
```

#### The correlation plot after feature creation elicits some new insights:

* Complaints per enplaned indiviudal have a strong positive correlation with the proportion of cancelled flights
* There is a strong negative relationship between the number of enplaned individuals per flight, and the number of complaints per enplaned individual
* There is a moderately strong relationship between the proportion of cancelled flights and the number of scheduled flights
* There is a strong positive relationship between the number of flights and the number of baggage complaints

## Exploratory Data Analysis

```{r}
# create time series plot
plot(american_eagle.ts, xlab = "Year", ylab = "Monthly Complaints", ylim = c(1000, 40000), col="green",lwd = 1.5,
     bty="l", main = "Baggage Complaints by Airline and Month")
lines(hawaiian.ts, col="orange", lwd = 1.5, bty="l")
lines(united.ts, col="steelblue", lwd = 1.5, bty="l")

# Add a legend
legend("topright", legend = c("United", "American Eagle", "Hawaiian"), col = c("steelblue", "green", "orange"), lwd = 1.5)
```

Throughout the dataset, United Airlines consistently receives the highest number of complaints regarding mishandled baggage each month, while Hawaiian Airlines consistently records the lowest number of complaints in every month.

The above plot shows a mild seasonality for American Eagle and United Airlines as there is a gradual increase in the baggage complaints at the beginning of each year. Additionally, it does not exhibit any trend as there is an increase and decrease in the baggage complaints for the three airlines.

### Monthly Average Complaints for each airline

```{r}
# Monthly average complaints for each airline
avg_complaints <- airlines_df %>% 
  group_by(Airline, Month) %>% summarise_at(vars(Baggage), list(Avg_Complaints = mean))

ggplot(avg_complaints, aes(x = Month, y = Avg_Complaints)) + 
  geom_line(aes(color = Airline, group = Airline)) + 
  labs(title = 'Monthly Average complaints', 
       x = 'Month', 
       y = 'Average No of complaints') 
```
Based on the above chart, on average,  United consistently receives a higher average number of complaints in comparison to other airlines. Additionally, it is observed that the number of complaints shows an increase from May to August and from November to December. Furthermore, the trend for Hawaiian Airlines appears comparatively stable, with less fluctuation compared to the other airlines.


### Proportion of complaints for each airline

```{r}
# Create a Date variable from the 'Date' column
airlines_df$Date <- as.Date(paste(airlines_df$Year, airlines_df$Month, "01", sep = "-"))

# Line plot for complaints for every month for each airline
ggplot(airlines_df, aes(x = Date, y = Complaints_per_enplaned, group = Airline, color = Airline)) +
  geom_line() +
  labs(title = "",
       x = "Date",
       y = "") +  
  scale_x_date(date_labels = "%Y", date_breaks = "2 years") +  
  theme_minimal() +
  facet_wrap(~Airline, scales = "free_y") +  # Separate plots for each airline
  theme(
    axis.title.y = element_blank(),  
    legend.position = "none",  
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
Upon examining the individual charts for each airline, it becomes evident that there was a rise in trends from 2004 to 2006, followed by a subsequent decline. Moreover, the plot highlights that Hawaiian Airlines had a big increase in complaints in 2010, while other airlines had a more steady rise in the number of complaints.

### Explore cancellations over the years

```{r}
# Explore cancellations over the years
ggplot(airlines_df, aes(x = as.Date(paste(Year, Month, "01", sep = "-")), y = Cancelled, group = Airline, color = Airline)) +
  geom_line() + 
  labs(title = "Cancellations Over the Years", 
       x = "Date", 
       y = "Number of Cancellations")
```
The above plot illustrates the annual trend of cancelled flights. It is observed that American Eagle has the highest count of cancelled flights and highest frequency of cancellations occur in January.

### Explore the trend in the Cancelled flights for each airline

```{r}
# Create a Date variable from the 'Date' column
airlines_df$Date <- as.Date(paste(airlines_df$Year, airlines_df$Month, "01", sep = "-"))

# Line plot of cancelled flights for each airline
ggplot(airlines_df, aes(x = Date, y = Cancelled, group = Airline, color = Airline)) +
  geom_line() +
  labs(title = "",
       x = "Date",
       y = "") +  
  scale_x_date(date_labels = "%Y", date_breaks = "2 years") +  
  theme_minimal() +
  facet_wrap(~Airline, scales = "free_y") +  # Separate plots for each airline
  theme(
    axis.title.y = element_blank(),  
    legend.position = "none",  
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
In the provided graph, American Eagle and United share a similar pattern in the number of cancelled flights over the years, while Hawaiian Airlines demonstrates relatively fewer ups and downs. Hawaiian flight cancellations reached their highest point in 2008 and started to decrease afterward.

### Explore monthly Cancelled proportions 

```{r}
# Working on this 

# ggplot(airlines_df, aes(x = Month, y = Cancelled_prop, color = Airline)) +
#   geom_line() +
#   theme_minimal() +
#   labs(title = "Cancelled Proportions of Each Airline Over Months",
#        x = "Month",
#        y = "") +
#   scale_color_discrete(name = "Airline") +
#   theme(legend.position = "top")
```

### Relationship between Baggage Complaints and Cancelled Flights

```{r}
# Calculate correlation coefficient
corr_coeff <- cor(airlines_df$Cancelled, airlines_df$Baggage)
cat("Correlation coefficient between Cancelled Flights and Baggage Complaints:", corr_coeff, "\n")

# Scatter plot of Baggage vs. Cancelled
ggplot(airlines_df, aes(x = Cancelled, y = Baggage, color=Airline)) +
  geom_point() +
  labs(title = "Scatter plot of Cancelled Flights vs. Baggage Complaints",
       x = "Cancelled Flights",
       y = "Baggage Complaints")
```
A correlation coefficient of 0.5944 indicates a moderate positive correlation between the number of cancelled flights and the number of baggage complaints. 

Based on the plot, it is observed that there is a general upward trend indicating that there is a tendency for higher baggage complaints when there are more cancelled flights. Moreover, it is important to note that correlation does not imply causation. While there is a statistical association between cancelled flights and baggage complaints, it doesn't necessarily mean that one causes the other. There could be other factors influencing both variables.


## Modeling

To being looking at airline baggage complaints, it is important to see the trends for each airline. One plot has been created for each airline, so we can clearly see the time series models for each airline. In addition to trends, a moving average has been added to aid in visualization of the data.

American Eagle Airlines Trend

```{r}
# Fit a trend line model
american_eagle_trend_model <- lm(american_eagle.ts ~ time(american_eagle.ts))

# Create a data frame for american eagle airlines
american_eagle_df <- data.frame(
  date = time(american_eagle.ts), 
  complaints = as.numeric(american_eagle.ts), 
  trend = predict(american_eagle_trend_model),
  moving_avg = ma(american_eagle.ts, order = 12))

colors <- c("American Eagle Airlines" = "green", "Trend" = "black", "Moving Average" = "red")

# Create a ggplot with trend line and legend
ggplot(american_eagle_df, aes(x = date)) +
  geom_line(aes(y = complaints, color = "American Eagle Airlines"), size = 1) +
  geom_line(aes(y = trend, color = "Trend"), size = 1) +
  geom_line(aes(y = moving_avg, color = "Moving Average"), size = 1) +
  labs(title = "Baggage Complaints for American Eagle Airlines",
       x = "Year", y = "Number of Complaints", color = "Legend") +
  scale_color_manual(name = "Legend", values = colors) +
  ylim(c(5000, 30000))
```

```{r}
# Get the trend for each season (month) for American Eagle Airlines
ae_lm_trend_season <- tslm(american_eagle.ts ~ trend + I(trend^2) + season)
ae_lm_trend_season
```

```{r}
# ACF plot of American Eagle Airlines
Acf(american_eagle.ts, lag.max = 12, main = "ACF Plot of American Airlines Baggage Complaints")
```

United Airlines Trend

```{r}
# Fit a trend line model
united_trend_model <- lm(united.ts ~ time(united.ts))

# Create a data frame for united airlines
united_df <- data.frame(
  date = time(united.ts), 
  complaints = as.numeric(united.ts), 
  trend = predict(united_trend_model),
  moving_avg = ma(united.ts, order = 12))

colors <- c("United Airlines" = "blue", "Trend" = "black", "Moving Average" = "red")

# Create a ggplot with trend line and legend
ggplot(united_df, aes(x = date)) +
  geom_line(aes(y = complaints, color = "United Airlines"), size = 1) +
  geom_line(aes(y = trend, color = "Trend"), size = 1) +
  geom_line(aes(y = moving_avg, color = "Moving Average"), size = 1) +
  labs(title = "Baggage Complaints for United Airlines",
       x = "Year", y = "Number of Complaints", color = "Legend") +
  scale_color_manual(name = "Legend", values = colors) +
  ylim(c(9000, 42000))
```

```{r}
# Get the trend for each season (month) for United Airlines
united_lm_trend_season <- tslm(united.ts ~ trend + I(trend^2) + season)
united_lm_trend_season
```

```{r}
# ACF plot of United Airlines
Acf(united.ts, lag.max = 12, main = "ACF Plot of United Airlines Baggage Complaints")
```

Hawaiian Airlines Trend

```{r}
# Fit a trend line model
hawaiian_trend_model <- lm(hawaiian.ts ~ time(hawaiian.ts))

# Create a data frame for ggplot
hawaiian_df <- data.frame(
  date = time(hawaiian.ts), 
  complaints = as.numeric(hawaiian.ts), 
  trend = predict(hawaiian_trend_model),
  moving_avg = ma(hawaiian.ts, order = 12))

colors <- c("Hawaiian Airlines" = "orange", "Trend" = "black", "Moving Average" = "red")

# Create a ggplot with trend line and legend
ggplot(hawaiian_df, aes(x = date)) +
  geom_line(aes(y = complaints, color = "Hawaiian Airlines"), size = 1) +
  geom_line(aes(y = trend, color = "Trend"), size = 1) +
  geom_line(aes(y = moving_avg, color = "Moving Average"), size = 1) +
  labs(
    title = "Baggage Complaints for Hawaiian Airlines",
    x = "Year", y = "Number of Complaints", color = "Legend"
  ) +
  scale_color_manual(name = "Legend", values = colors) +
  ylim(c(1000, 3000))
```

```{r}
# Get the trend for each season (month) for Hawaiian Airlines
hawiian_lm_trend_season <- tslm(hawaiian.ts ~ trend + I(trend^2) + season)
hawiian_lm_trend_season
```

```{r}
# ACF plot of Hawaiian Airlines
Acf(hawaiian.ts, lag.max = 12, main = "ACF Plot of Hawiian Airlines Baggage Complaints")
```

Based on trends from all three airlines, American Eagle and United are trending towards a decline in the number of total complains. However, Hawaiian airlines is trending towards an increase in total baggage complaints. American Eagle Airlines is pushing above the moving average, United Airlines is below the moving average, but is starting to experience an upward shift, and Hawaiian Airlines is pushing far above the moving average and shows no sign of slowing down.

Split the data into train and validation datasets.

```{r}
# Set the cutoff date for the training period
cutoff <- as.Date("2010-01-01")

# Create the training and validation datasets
ae_train_data <- subset(airlines_df, Airline == "American Eagle" & Date < cutoff)
ae_validation_data <- subset(airlines_df, Airline == "American Eagle" & Date >= cutoff)

# Create a time series for training data
ae_ts_train <- ts(ae_train_data$Baggage, 
                  start = c(2004, 1), end = c(2009, 12), frequency = 12)

# Create a time series for validation data
ae_ts_valid <- ts(ae_validation_data$Baggage, 
                  start = c(2010, 1), end = c(2010, 12), frequency = 12)
```

```{r}
# Create an ARIMA model for American Eagle Airlines
arima_american_eagle <- auto.arima(ae_ts_train)  
pred_american_eagle <- forecast(arima_american_eagle, h = 12)
pred_american_eagle_fitted <- pred_american_eagle$fitted

# create a new data frame with the original and ARIMA model
residuals_df_american_eagle <- data.frame(
  time = time(ae_ts_train),
  original = as.numeric(ae_ts_train),
  smoothed = pred_american_eagle_fitted
)

colors <- c("American Eagle Airlines" = "green", "American Eagle Airlines ARIMA" = "chartreuse4")

# Create ggplot
ggplot(residuals_df_american_eagle, aes(x = time)) +
  geom_line(aes(y = original, color = "American Eagle Airlines"), size = 1) +
  geom_line(aes(y = smoothed, color = "American Eagle Airlines ARIMA"), size = 1) +
  labs(
    title = "American Eagle Airlines Baggage Complaints",
    x = "Year",
    y = "Number of Complaints"
  ) +
  scale_color_manual(name = "Legend", values = colors) +
  ylim(c(5000, 30000))
```

```{r}
# Calculate the mean of the training data
mean_american_eagle <- mean(as.numeric(ae_ts_train))

# Calculate the mean for the length of the training dataset
american_eagle_predicted_values <- rep(mean_american_eagle, length(ae_ts_train))

# Compute RMSE
american_eagle_rmse_pred <- sqrt(mean((american_eagle_predicted_values - as.numeric(ae_ts_train))^2))
print(paste("RMSE for the training period (using mean):", round(american_eagle_rmse_pred, 2)))
```

```{r}
# Get the summary of the ARIMA model (RMSE, forecasts, etc.)
summary(pred_american_eagle)
```

```{r}
# Create the training and validation datasets
united_train_data <- subset(airlines_df, Airline == "United" & Date < cutoff)
united_validation_data <- subset(airlines_df, Airline == "United" & Date >= cutoff)

# Create a time series for training data
united_ts_train <- ts(united_train_data$Baggage, 
                  start = c(2004, 1), end = c(2009, 12), frequency = 12)

# Create a time series for validation data
united_ts_valid <- ts(united_validation_data$Baggage, 
                  start = c(2010, 1), end = c(2010, 12), frequency = 12)
```


```{r}
# Create an ARIMA model for Hawaiian Airlines
arima_united <- auto.arima(united_ts_train)  
pred_united <- forecast(arima_united, h = 12)
pred_united_fitted <- pred_united$fitted

# Data preparation
residuals_df_american_eagle <- data.frame(
  time = time(united_ts_train),
  original = as.numeric(united_ts_train),
  smoothed = pred_united_fitted
)

colors <- c("United Airlines" = "blue", "United Airlines ARIMA" = "deepskyblue")

# Create ggplot
ggplot(residuals_df_american_eagle, aes(x = time)) +
  geom_line(aes(y = original, color = "United Airlines"), size = 1) +
  geom_line(aes(y = smoothed, color = "United Airlines ARIMA"), size = 1) +
  labs(
    title = "United Airlines Baggage Complaints",
    x = "Year",
    y = "Number of Complaints"
  ) +
  scale_color_manual(name = "Legend", values = colors) +
  ylim(c(9000, 42000))
```

```{r}
# Calculate the mean of the training data
mean_united <- mean(as.numeric(united_ts_train))

# Calculate the mean for the length of the training dataset
united_predicted_values <- rep(mean_united, length(united_ts_train))

# Compute RMSE
hawiian_rmse_pred <- sqrt(mean((united_predicted_values - as.numeric(united_ts_train))^2))
print(paste("RMSE for the training period (using mean):", round(hawiian_rmse_pred, 2)))
```


```{r}
# Get the summary of the ARIMA model (RMSE, forecasts, etc.)
summary(pred_united)
```

```{r}
# Create the training and validation datasets
hawaiian_train_data <- subset(airlines_df, Airline == "Hawaiian" & Date < cutoff)
hawaiian_validation_data <- subset(airlines_df, Airline == "Hawaiian" & Date >= cutoff)

# Create a time series for training data
hawaiian_ts_train <- ts(hawaiian_train_data$Baggage, 
                  start = c(2004, 1), end = c(2009, 12), frequency = 12)

# Create a time series for validation data
hawaiian_ts_valid <- ts(hawaiian_validation_data$Baggage, 
                  start = c(2010, 1), end = c(2010, 12), frequency = 12)
```


```{r}
# Create an ARIMA model for Hawaiian Airlines
arima_hawaiian <- auto.arima(hawaiian_ts_train)  
pred_hawaiian <- forecast(arima_hawaiian, h = 12)
pred_hawaiian_fitted <- pred_hawaiian$fitted

# Data preparation
residuals_df_american_eagle <- data.frame(
  time = time(hawaiian_ts_train),
  original = as.numeric(hawaiian_ts_train),
  smoothed = pred_hawaiian_fitted
)

colors <- c("Hawaiian Airlines" = "orange", "Hawaiian Airlines ARIMA" = "chocolate")

# Create ggplot
ggplot(residuals_df_american_eagle, aes(x = time)) +
  geom_line(aes(y = original, color = "Hawaiian Airlines"), size = 1) +
  geom_line(aes(y = smoothed, color = "Hawaiian Airlines ARIMA"), size = 1) +
  labs(
    title = "Hawaiian Airlines Baggage Complaints",
    x = "Time",
    y = "Number of Complaints"
  ) +
  scale_color_manual(name = "Legend", values = colors) +
  ylim(c(1000, 3000))
```

```{r}
# Calculate the mean of the training data
mean_hawiian <- mean(as.numeric(hawaiian_ts_train))

# Calculate the mean for the length of the training dataset
hawiian_predicted_values <- rep(mean_hawiian, length(hawaiian_ts_train))

# Compute RMSE
hawiian_rmse_pred <- sqrt(mean((hawiian_predicted_values - as.numeric(hawaiian_ts_train))^2))
print(paste("RMSE for the original data (using mean):", round(hawiian_rmse_pred, 2)))
```


```{r}
# Get the summary of the ARIMA model (RMSE, forecasts, etc.)
summary(pred_hawaiian)
```
